Haskell for Logic Programmers:\\ Compiling Functional Logic Programs to Haskell\thanks{ A slightly modified version of this paper will appear in Proc.\ of WFLP 2011.}

### 2011-06-14 papers/ciclops2011/paper.tex
CICLOPS'11 paper added

\title{KiCS2: A New Compiler from Curry to Haskell}

@@

\begin{abstract}
In this paper we present our first steps towards a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic results as values of the data types corresponding
to the results.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperform other existing implementations of Curry.
\end{abstract}

@@

Functional logic languages integrate the most important
features of functional and logic languages
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they combine higher-order functions and demand-driven
evaluation from functional programming with logic programming features
like non-deterministic search and computing with partial information
(logic variables).
The combination of these features
has led to new design patterns \cite{AntoyHanus02FLOPS}
and better abstractions for application programming,
e.g., as shown for programming with databases
\cite{BrasselHanusMueller08PADL,Fischer05},
GUI programming \cite{Hanus00PADL},
web programming \cite{Hanus01PADL,Hanus06PPDP,HanusKoschnicke10PADL},
or string parsing \cite{CaballeroLopez99}.

The implementation of functional logic languages is challenging
since a reasonable implementation has to support the operational
features mentioned above.
One possible approach
is the design of new abstract machines appropriately supporting
these operational features and implementing them in some (typically,
imperative) language, like C \cite{Lux99FLOPS}
or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
Another approach is the reuse of already existing implementations
of some of these features by translating
functional logic programs into either logic or functional languages.
For instance, if one compiles into Prolog, one can reuse
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information. However, one has to implement demand-driven evaluation
and higher-order functions \cite{AntoyHanus00FROCOS}.
A disadvantage of this approach is the commitment to a fixed
search strategy (backtracking).

@@

Nevertheless, the translation of functional logic languages
into other high-level languages is attractive:
it limits the implementation efforts compared to an implementation
from scratch and one can exploit the existing implementation technologies,
provided that the efforts to implement the missing features
are reasonable.

In this paper we describe an implementation that is based
on the latter principle. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs based on the ideas shown
in \cite{BrasselFischer08IFL}.
The difficulty of such an implementation is the fact
that non-deterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
as the outcome of any operation could be potentially non-deterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
non-deterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
This idea is also the basis of the Curry implementation KiCS
\cite{BrasselHuch07,BrasselHuch09}.
However, KiCS is based on unsafe features of Haskell
that inhibit the use of optimizations provided by Haskell compilers
like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
In contrast, our implementation, called KiCS2, avoids such unsafe features.
In addition, we also support more flexible search strategies
and new features to encapsulate non-deterministic computations
(which are not described in detail in this paper due to lack of space).

The general objective of our approach is the support
of flexible strategies to explore the search space
resulting from non-deterministic computations.
In contrast to Prolog-based implementations
that use backtracking and, therefore, are incomplete,
we also want to support complete strategies like breadth-first search,
iterative deepening or parallel search (in order to exploit
multi-core architectures). We achieve this goal by an
explicit representation of the search space as data
that can be traversed by various operations.
Moreover, purely deterministic computations
are implemented as purely functional programs so that
they are executed with almost the same efficiency
as their purely functional counterparts.

@@

The syntax of the functional logic language Curry \cite{Hanus06Curry}
is close to Haskell \cite{PeytonJones03Haskell}.
In addition, Curry allows free (logic)
variables in conditions and right-hand sides of defining rules.
In contrast to functional programming and similarly to logic programming,
operations can be defined by overlapping rules so that
they might yield more than one result on the same input.
Such operations are also called \emph{non-deterministic}.
For instance, Curry offers a \emph{choice} operation that is predefined by
the following rules:
%
\begin{curry}
  x ? _ = x
  _ ? y = y
\end{curry}
%
Thus, we can define a non-deterministic operation \code{aBool} by
\label{ex:aBool}
%
\begin{curry}
  aBool = True ? False
\end{curry}
%
so that the expression \ccode{aBool} has two values:
\code{True} and \code{False}.

@@

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each non-deterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a non-deterministic argument
yields non-deterministic results for each of the argument's values.
\item Each choice introduced in a computation is supplied with its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques developed for purely functional programs
and make the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we avoid unsafe features in our implementation.
Instead, we thread some global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program, i.e., a Curry operation of type
\begin{curry}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$->$~\tau$
\end{curry}
is translated into a Haskell function of type
\begin{haskell}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$-> IDSupply ->$~\tau$
\end{haskell}
%
Conceptually, one can consider \code{IDSupply}
as an infinite set of identifiers that is created at the beginning
of an evaluation by the operation \code{initSupply}.
The operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
The split operations \code{leftSupply} and \code{rightSupply}
are used when an operation calls two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the called operations must be supplied with their individual disjoint
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main :: Bool
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main :: IDSupply -> Bool
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
The type \code{IDSupply} can be implemented in various ways.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

Functional logic computations can also fail, e.g.,
due to partially defined operations.
Computing with failures is a typical programming technique
and provides for specific programming patterns \cite{AntoyHanus02FLOPS}.
Hence, in contrast to functional programming,
a failing computation should not abort the complete evaluation
but it should be considered as some part of a computation that
does not produce a meaningful result.
In order to implement this behavior, we extend
each data type by a further constructor \code{Fail}
and complete each operation containing matching rules
by a final rule that matches everything and returns the value \code{Fail}.
For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is extended as follows:\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
Note that the final rule returns \code{Fail} if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation, i.e., it also propagates failures.

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation supports various options together with different
implementations of \code{IDSupply}.
For instance, a simple but efficient implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
Consequently, the implementation of \code{IDSupply} requires
an infinite set of memory cells which can represented as a tree structure:
\begin{haskell}
  data IDSupply = IDSupply ID IDSupply IDSupply
  thisID      (IDSupply r _ _) = r
  leftSupply  (IDSupply _ s _) = s
  rightSupply (IDSupply _ _ s) = s
  \end{haskell}
The infinite tree of memory cells (with initial value \code{NoChoice})
can be constructed as follows,
where \code{unsafeInterleaveIO} is used to construct
the tree on demand:
\begin{haskell}
  initSupply = getIDTree$\listline$
  getIDTree = do s1 <- unsafeInterleaveIO getIDTree
                 s2 <- unsafeInterleaveIO getIDTree
                 r  <- unsafeInterleaveIO (newIORef NoChoice)
                 return (IDSupply (ID r) s1 s2)
  \end{haskell}
Using memory cells,
the implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

Of course, printing all values via depth-first search
is only one option which is not sufficient in case of infinite search
spaces. For instance, one can easily define an operation
that prints only the first solution. Due to the lazy evaluation
strategy of Haskell, such an operation can also be applied to
infinite choice structures.
In order to abstract from these different printing options,
our implementation contains a more general
approach by translating choice structures into monadic structures
w.r.t.\ various strategies (depth-first search, breadth-first search,
iterative deepening, parallel search).
This allows for an independent processing of the resulting monadic structures,
e.g.,
by an interactive loop where the user can request the individual values.

@@

Due to the fact that we represent non-deterministic results
in a data structure rather than as a computation as in implementations
based on backtracking, we can provide different methods to
explore the search space containing the different result values.
We have already seen in Section~\ref{sec:printValsDFS}
how this search space can be explored to print all values in
depth-first order.
Apart from this simple approach,
our implementation contains various strategies
(depth-first, breadth-first, iterative deepening, parallel search)
to transform a choice structure into a list of results
that can be printed in different ways (e.g., all solutions, only the first
solution, or one after another by user requests).
Actually, the user can set options to select the search strategy and
the printing method.

@@

In addition to the different search strategies to evaluate
top-level expressions, our system also contains a primitive operation
\begin{curry}
 searchTree :: a -> IO (SearchTree a)
\end{curry}
to translate the search space caused by the evaluation of its argument
into a tree structure of the form
\begin{curry}
 data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
With this primitive,
the programmer can define its own search strategy or
collect all non-deterministic values into a list structure
for further processing \cite{BrasselHuch07}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Related Work}
\label{sec:Conclusions}

We have presented a new system to compile functional logic programs
into purely functional programs.
In order to be consistent with the call-time choice semantics
of functional logic languages like Curry or TOY,
we represent non-deterministic values
in choice structures where each choice has an identification.
Values for such choice identifiers are passed through
non-deterministic operations so that fresh identifiers are available
when a new choice needs to be created.
The theoretical justification of this implementation technique
is provided in \cite{Brassel11Thesis}.
Apart from the parser, where we reused an existing one
implemented in Haskell, the compiler is completely written in Curry.

Due to the representation of non-deterministic values as data,
our system easily supports various search strategies
in constrast to Prolog-based implementations.
Since we compile Curry programs into Haskell,
we can exploit the implementation efforts done
for functional programming.
Hence, purely functional parts of functional logic programs
can be executed with almost the same efficiency as Haskell programs.
Our benchmarks show that even the execution of the
non-deterministic parts can compete with other
implementations of Curry.

@@

If non-deterministic results are collected in data structures,
one has more fine-grained control over non-deterministic steps.
For instance, \cite{AlqaddoumiAntoyFischerReck10} proposes
pull-tab steps to move non-determinism from arguments to
the result position of a function. Antoy \cite{Antoy11ICLP}
shows that single pull-tab steps are semantics-preserving.
Thus, it is not necessary to move each choice to the root
of an expression, as done in our implementation,
but one could also perform further local computations in the arguments
of a choice before moving it up. This might be a reasonable strategy
if all non-deterministic values are required but many computations fail.
However, the general effects of such refinements need further
investigations.

@@

\begin{thebibliography}{10}

@@

\bibitem{Antoy11ICLP}
S.~Antoy.
\newblock On the correctness of the pull-tab transformation.
\newblock In {\em To appear in Proceedings of the 27th International Conference
  on Logic Programming (ICLP 2011)}, 2011.

@@

@@

\bibitem{Zartmann97SAS}
F.~Zartmann.
\newblock Denotational Abstract Interpretation of Functional Logic Programs.
\newblock In {\em Proc. of the 4th International Symposium on Static Analysis
  (SAS'97)}, pp. 141--156. Springer LNCS 1302, 1997.

>>>>

\title{Haskell for Logic Programmers:\\
 Compiling Functional Logic Programs to Haskell\thanks{%
A slightly modified version of this paper will appear in
Proc.\ of WFLP 2011.}}

@@

\begin{abstract}
In this paper we present recent results about a new system to compile
functional logic programs of the source language Curry into purely
functional Haskell programs. The implementation is based on the idea
to represent non-determinism as data in Haskell. This enables the
application of various search strategies to extract values from the
search space. We show by several benchmarks that our implementation
can compete with or outperforms other existing implementations of
Curry.
\end{abstract}

@@

Functional logic languages integrate the most important
features of functional and logic languages
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they combine higher-order functions and demand-driven
evaluation from functional programming with logic programming features
like non-deterministic search and computing with partial information
(logic variables).
The combination of these features
has led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
and better abstractions for application programming,
e.g., as shown for programming with databases
\cite{BrasselHanusMueller08PADL,Fischer05},
GUI programming \cite{Hanus00PADL},
web programming \cite{Hanus01PADL,Hanus06PPDP,HanusKoschnicke10PADL},
or string parsing \cite{CaballeroLopez99}.

The implementation of functional logic languages is challenging
since the operational features mentioned above have to be supported.
One possible approach
is the design of abstract machines appropriately supporting
these operational features and implementing them in some (typically,
imperative) language, like C \cite{Lux99FLOPS}
or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
Another approach is the reuse of already existing implementations
of some of these features by translating
functional logic programs into either logic or functional languages.
For instance, if one compiles into Prolog, one can reuse
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information. However, one has to implement demand-driven evaluation
and higher-order functions \cite{AntoyHanus00FROCOS}.
A disadvantage of this approach is the commitment to a fixed
search strategy (backtracking).

@@

In this paper we describe an implementation that is based
on the latter principle. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs based on the ideas shown
in \cite{BrasselFischer08IFL}.
The difficulty of such an implementation is the fact
that non-deterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
as the outcome of any operation could be potentially non-deterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
non-deterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
Hence, each expression evaluates into an explicit representation
of its search space. Therefore, our implementation is not
committed to backtracking (in contrast to Prolog-based implementations)
but also supports complete strategies like breadth-first search,
iterative deepening or parallel search (in order to exploit
multi-core architectures).
Moreover, purely deterministic computations
are implemented as purely functional programs so that
they are executed with almost the same efficiency
as their purely functional counterparts.

@@

The syntax of the functional logic language Curry \cite{Hanus06Curry}
is close to Haskell \cite{PeytonJones03Haskell}, i.e.,
(type) variables and function names usually
start with lowercase letters and the names of type and data constructors
start with an uppercase letter. The application of $f$
to $e$ is denoted by juxtaposition (``$f~e$'').
In addition, Curry allows free (logic)
variables in conditions and right-hand sides of defining rules.
In contrast to functional programming and similarly to logic programming,
operations can be defined by overlapping rules so that
they might yield more than one result on the same input.
Such operations are also called \emph{non-deterministic}.
For instance, Curry offers a \emph{choice} operation that is predefined by
the following rules:
%
\begin{curry}
  x ? _ = x
  _ ? y = y
\end{curry}
%
Thus, we can define a non-deterministic operation \code{aBool} by
\label{ex:aBool}
%
\begin{curry}
  aBool = True ? False
\end{curry}
%
so that the expression \ccode{aBool} has two values:
\code{True} and \code{False}.

@@

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each non-deterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a non-deterministic argument
yields non-deterministic results for each of the argument's values.
\item Each choice introduced in a computation is supplied with its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques developed for purely functional programs
and make the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we avoid unsafe features in our implementation.
Instead, we thread some global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program, i.e., a Curry operation of type
\begin{curry}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$->$~\tau$
\end{curry}
is translated into a Haskell function of type
\begin{haskell}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$-> IDSupply ->$~\tau$
\end{haskell}
%
Conceptually, one can consider \code{IDSupply}
as an infinite set of identifiers that is created at the beginning
of an evaluation by the operation \code{initSupply}.
The operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
The split operations \code{leftSupply} and \code{rightSupply}
are used when an operation calls two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the called operations must be supplied with their individual disjoint
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main :: Bool
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main :: IDSupply -> Bool
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
The type \code{IDSupply} can be implemented in various ways.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compiler is parameterized over different
implementations of \code{IDSupply}
in order to choose the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply} and $o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$,
then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

Similarly to logic programming,
functional logic computations can also fail, e.g.,
due to partially defined operations.
Computing with failures is a typical programming technique
and provides for specific programming patterns \cite{AntoyHanus02FLOPS}.
Hence, in contrast to functional programming,
a failing computation should not abort the complete evaluation
but it should be considered as some part of a computation that
does not produce a meaningful result.
In order to implement this behavior, we extend
each data type by a further constructor \code{Fail}
and complete each operation containing matching rules
by a final rule that matches everything and returns the value \code{Fail}.
For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is extended as follows:\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
Note that the final rule returns \code{Fail} if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation, i.e., it also propagates failures.

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a potentially
non-deterministic computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation supports various options together with different
implementations of \code{IDSupply}.
For instance, a simple but efficient implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
Consequently, the implementation of \code{IDSupply} requires
an infinite set of memory cells which can represented as a tree structure:
\begin{haskell}
  data IDSupply = IDSupply ID IDSupply IDSupply
  thisID      (IDSupply r _ _) = r
  leftSupply  (IDSupply _ s _) = s
  rightSupply (IDSupply _ _ s) = s
  \end{haskell}
The infinite tree of memory cells (with initial value \code{NoChoice})
can be constructed as follows,
where \code{unsafeInterleaveIO} is used to construct
the tree on demand:
\begin{haskell}
  initSupply = getIDTree$\listline$
  getIDTree = do s1 <- unsafeInterleaveIO getIDTree
                 s2 <- unsafeInterleaveIO getIDTree
                 r  <- unsafeInterleaveIO (newIORef NoChoice)
                 return (IDSupply (ID r) s1 s2)
  \end{haskell}
Using memory cells,
the implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

Of course, printing all values via depth-first search
is only one option which is not sufficient in case of infinite search
spaces. For instance, one can easily define an operation
that prints only the first solution. Due to the lazy evaluation
strategy of Haskell, such an operation can also be applied to
infinite choice structures.
In order to abstract from these different printing options,
our implementation contains a more general
approach by translating choice structures into monadic structures
w.r.t.\ various strategies (depth-first search, breadth-first search,
iterative deepening, parallel search).
This allows for an independent processing of the resulting monadic structures,
e.g.,
by an interactive loop where the user can request the individual values.
Therefore, our implementation separates the logic
(what are the values of an expression?) from the control
(in which order should the values be computed?),
in contrast to Prolog-based implementations,
like PAKCS \cite{Hanus10PAKCS} or TOY \cite{Lopez-FraguasSanchez-Hernandez99},
that are committed to a fixed search strategy (backtracking).

@@

Due to the fact that we represent non-deterministic results
in a data structure rather than as a computation as in implementations
based on backtracking, we can provide various methods to
explore the search space containing the different result values.
We have already seen in Section~\ref{sec:printValsDFS}
how this search space can be explored to print all values in
depth-first order.
Apart from this simple approach,
our implementation contains various strategies
(depth-first, breadth-first, iterative deepening, parallel search)
to transform a choice structure into a list of results
that can be printed in different ways (e.g., all solutions, only the first
solution, or one after another by user requests).
Actually, the user can set options to select the search strategy and
the printing method.

@@

In addition to the different search strategies to evaluate
top-level expressions, our system also contains a primitive operation
\begin{curry}
 getSearchTree :: a -> IO (SearchTree a)
\end{curry}
to translate the search space caused by the evaluation of its argument
into a tree structure of the form
\begin{curry}
 data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
With this primitive,
the programmer can define its own search strategy or
collect all non-deterministic values into a list structure
for further processing \cite{BrasselHuch07}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:Conclusions}

We have presented a new system to compile functional logic programs
into purely functional programs.
In order to be consistent with the call-time choice semantics
of functional logic languages like Curry or TOY,
we represent non-deterministic values
in choice structures where each choice has an identification.
Values for such choice identifiers are passed through
non-deterministic operations so that fresh identifiers are available
when a new choice needs to be created.
The theoretical justification of this implementation technique
is provided in \cite{Brassel11Thesis}.

Due to the representation of non-deterministic values as data,
our system easily supports various search strategies
in contrast to Prolog-based implementations.
Actually, our implementation can also be used as a testbed
to explore different strategies for non-deterministic computations.
For instance, \cite{AlqaddoumiAntoyFischerReck10} proposes
pull-tab steps to move non-determinism from arguments to
the result position of a function. Antoy \cite{Antoy11ICLP}
shows that single pull-tab steps are semantics-preserving.
Thus, it is not necessary to move each choice to the root
of an expression, as done in our current implementation,
but one could also perform further local computations in the arguments
of a choice before moving it up. This might be a reasonable strategy
if all non-deterministic values are required but many computations fail.
However, the general effects of such refinements need further
investigation that can be performed by slightly changing our
compilation scheme.

Since we compile Curry programs into Haskell,
we can exploit the implementation efforts done
for functional programming.
Hence, purely functional parts of functional logic programs
are executed with almost the same efficiency as Haskell programs.
Our benchmarks show that even the execution of the
non-deterministic parts can compete with other
implementations of Curry.
Apart from the parser, where we reused an existing one
implemented in Haskell, our compiler is completely written in Curry.
The first version of the compiler was implemented
with the Prolog-based Curry implementation PAKCS \cite{Hanus10PAKCS}.
After bootstrapping this compiler, it became approximately 20 times
faster. This shows that the results of the small benchmarks
presented in this paper are also valid for larger applications.

@@

@@

%\bibliographystyle{plain}
%\bibliography{mh}
\begin{thebibliography}{10}

@@

\bibitem{Antoy11ICLP}
S.~Antoy.
\newblock On the correctness of pull-tabbing.
\newblock In {\em To appear in Proceedings of the 27th International Conference
  on Logic Programming (ICLP 2011)}, 2011.

@@

\bibitem{AntoyHanus11WFLP}
S.~Antoy and M.~Hanus.
\newblock New functional logic design patterns.
\newblock In {\em Proc.\ of the 20th International Workshop on Functional and
  (Constraint) Logic Programming (WFLP 2011)}. Springer LNCS (to appear), 2011.

@@

\bibitem{Zartmann97SAS}
F.~Zartmann.
\newblock Denotational abstract interpretation of functional logic programs.
\newblock In {\em Proc. of the 4th International Symposium on Static Analysis
  (SAS'97)}, pages 141--156. Springer LNCS 1302, 1997.
